\documentclass[a4paper,10pt]{article}
\usepackage{amssymb,amsmath ,amsthm, graphicx}
\usepackage[left=2cm,top=2cm,right=2cm,bottom=2cm,nohead,nofoot]{geometry}
\usepackage{enumitem}

\input{preamble.tex}
\usepackage{csquotes}
\usepackage{booktabs} 
% \usepackage{paralist} % for compactenum

%kyle's macros

\DeclareMathOperator{\tw}{tw}

\providecommand{\vvk}[2]{\vecn{#1}^{(#2)}}
\providecommand{\hvf}{\hvv{f}}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{proposition}{Proposition}


%% MAIN TEXT
\title{Network analysis preliminaries:\\
Graph theory and linear algebra}
\author{Kyle Kloster}


\begin{document}
\maketitle

\abstract{
Basic graph and matrix definitions, ideas, tools and facts.
 }


 %%%%%%%
 %%%%%%%
 %%%%      SECTION HEADING
 %%%%%%%
 %%%%%%%
\section{Overview}
This document summarizes the relevant terminology, ideas, and tools to dig into some network analysis. We're particularly interested in graph structures and sparsity (like cliques and k-cores), and their relationship to graph-matrix objects (like functions of the adjacency matrix and eigenvalues of the Laplacian). We'll define and discuss these terms (and many more) in the sections to come.

As with many subjects there is a wealth of terminology to cover at the start -- if it feels overwhelming, or unmotivated, don't worry. It is abstract and there is a bit much of it, but as we use the definitions and notation more and more, it will become easier. All of the terms here will be re-used a lot as we study and discuss objects and problems with neat theory and applications.

\input{prelim-notation.tex}

%%%%%%%
%%%%%%%
%%%%      SECTION HEADING
%%%%%%%
%%%%%%%
\section{Network Analysis Fundamentals}\label{sec:fundamentals}


\input{sec-eigen.tex}

%\input{sec-grapheigen.tex}

\input{sec-walks.tex}


\section{More on structured sparsity}\label{sec:structured-sparsity}
\input{sec-treedecomp.tex}

\input{sec-minors.tex}


\section{Eigen-fundamentals, Part 2}\label{sec:fundamentals:eigen2}

Don't worry about this section for a while -- this section will come in handy when we start trying to *prove* things about eigenvalues of certain graphs.

\input{sec-eigen2.tex}

%%%%%%%
%%%%%%%
%%%%      SECTION HEADING
%%%%%%%
%%%%%%%





{\footnotesize
\bibliographystyle{amsplain}
\bibliography{all-bibliography}
}

\end{document}
